---
title: "DTM Evaluation — `r params$site` `r gsub('_', '-', params$date)`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: kable
params:
  site:     "rr"
  date:     "2022_08_10"
  eval_dir: "lidar/output/rr_2022_08_10"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
library(here)
setwd(here::here())

summary_csv <- file.path(params$eval_dir, "dtm_eval_summary.csv")
if (!file.exists(summary_csv)) {
   stop("Summary CSV not found: ", summary_csv,
        "\nRun lidar/03_evaluate_dtm.R first.")
}
all_summary <- read.csv(summary_csv, stringsAsFactors = FALSE)

# DTM stems (unique, excluding the "overall" bookkeeping row).
dtm_stems <- unique(all_summary$dtm)

# Columns: keep pct_within_* whatever their exact names are.
pct_cols  <- grep("^pct_within", colnames(all_summary), value = TRUE)

# Plot names produced by 03_evaluate_dtm.R.
plot_names <- c("pred_vs_obs", "residual_hist", "residual_map",
                "residual_vs_elevation", "qq_residuals")
plot_labels <- c(
   pred_vs_obs           = "Predicted vs observed",
   residual_hist         = "Residual histogram",
   residual_map          = "Residual map",
   residual_vs_elevation = "Residual vs elevation",
   qq_residuals          = "Q-Q plot"
)

# Bias significance flag.
overall_rows <- all_summary[all_summary$group == "overall", ]
flagged_bias <- overall_rows[!is.na(overall_rows$bias_p) &
                               overall_rows$bias_p < 0.05, "dtm"]

# Helper: pretty-round a numeric column.
rnd <- function(x, d = 3) round(as.numeric(x), d)
```

## Overview

```{r overview}
# Count ECPs from the summary (use n column for overall row of first DTM).
n_ecp_overall <- overall_rows$n[1]
n_dtms        <- length(dtm_stems)

cat("**Site:**", params$site, " | ",
    "**Date:**", gsub("_", "-", params$date), " | ",
    "**DTMs evaluated:**", n_dtms, " | ",
    "**ECPs (overall):**", n_ecp_overall)
```

```{r ecp_class_counts}
# ECP counts by class (from first DTM's non-overall rows).
class_rows <- all_summary[all_summary$dtm == dtm_stems[1] &
                             all_summary$group != "overall", ]
class_tbl  <- class_rows[, c("group", "n")]
colnames(class_tbl) <- c("ECP class", "n")
knitr::kable(class_tbl[order(-class_tbl$n), ],
             row.names = FALSE,
             caption = "ECP counts by class (Logger Array excluded).")
```

```{r bias_flag, results = "asis"}
if (length(flagged_bias) > 0) {
   cat("> **Note:** The following DTMs have a mean vertical bias",
       "statistically distinguishable from zero (t-test p < 0.05):",
       paste(flagged_bias, collapse = ", "),
       ". Check the residual plots for systematic patterns.\n")
} else {
   cat("> Mean vertical bias is not statistically distinguishable",
       "from zero in any DTM (t-test p ≥ 0.05 for all).\n")
}
```

---

## Rankings

Bare-class RMSE (Berm + Training) reflects CSF ground-classification
performance on surfaces where the lidar can actually see the floor.
All-class RMSE is biased upward by tall-vegetation classes where
unfiltered canopy returns inflate the apparent error.

### Bare-class RMSE

```{r bare_rank}
bare_types <- c("Berm", "Training")
bare_rows  <- all_summary[all_summary$group %in% bare_types, ]

if (nrow(bare_rows) > 0) {
   bare_rank <- aggregate(
      rmse ~ dtm, data = bare_rows,
      FUN = function(x) round(mean(x, na.rm = TRUE), 3)
   )
   bare_rank <- bare_rank[order(bare_rank$rmse), ]
   colnames(bare_rank) <- c("DTM", "Mean RMSE (m)")
   knitr::kable(bare_rank, row.names = FALSE)
} else {
   cat("_No Berm or Training rows found._")
}
```

### Overall RMSE

```{r overall_rank}
show_cols <- c("dtm", "n", "mean_bias", "rmse", "mae", pct_cols)
show_cols <- show_cols[show_cols %in% colnames(overall_rows)]
rank_tbl  <- overall_rows[order(overall_rows$rmse), show_cols]
rank_tbl$mean_bias <- rnd(rank_tbl$mean_bias, 3)
rank_tbl$rmse      <- rnd(rank_tbl$rmse, 3)
rank_tbl$mae       <- rnd(rank_tbl$mae, 3)
for (pc in pct_cols) {
   rank_tbl[[pc]] <- rnd(rank_tbl[[pc]], 1)
}
colnames(rank_tbl)[colnames(rank_tbl) == "dtm"]       <- "DTM"
colnames(rank_tbl)[colnames(rank_tbl) == "mean_bias"] <- "Bias (m)"
colnames(rank_tbl)[colnames(rank_tbl) == "rmse"]      <- "RMSE (m)"
colnames(rank_tbl)[colnames(rank_tbl) == "mae"]       <- "MAE (m)"
knitr::kable(rank_tbl, row.names = FALSE)
```

---

## Full metrics table

All metric columns, all DTMs and classes.

```{r full_table}
show <- all_summary
show$mean_bias    <- rnd(show$mean_bias, 3)
show$median_bias  <- rnd(show$median_bias, 3)
show$rmse         <- rnd(show$rmse, 3)
show$mae          <- rnd(show$mae, 3)
show$mad          <- rnd(show$mad, 3)
show$max_abs      <- rnd(show$max_abs, 3)
show$q25_abs      <- rnd(show$q25_abs, 3)
show$q75_abs      <- rnd(show$q75_abs, 3)
show$q95_abs      <- rnd(show$q95_abs, 3)
show$slope        <- rnd(show$slope, 3)
show$bias_p       <- rnd(show$bias_p, 3)
for (pc in pct_cols) show[[pc]] <- rnd(show[[pc]], 1)
knitr::kable(show, row.names = FALSE)
```

---

## Plots {.tabset}

Five diagnostic plots per DTM.
**Q-Q plot:** a straight line means residuals are roughly normal,
which validates the t-test used for `bias_p`; heavy tails suggest
using a sign test instead.

```{r plots, results = "asis"}
for (stem in dtm_stems) {
   cat("\n### ", stem, "\n\n", sep = "")
   for (pname in plot_names) {
      png_path <- file.path(params$eval_dir,
                            paste0(stem, "_", pname, ".png"))
      if (file.exists(png_path)) {
         label <- if (pname %in% names(plot_labels)) {
            plot_labels[pname]
         } else {
            pname
         }
         cat("**", label, "**\n\n", sep = "")
         cat("![](",
             knitr::image_uri(png_path),
             "){width=80%}\n\n",
             sep = "")
      }
   }
}
```
