---
title: "DTM Evaluation — `r params$site` `r gsub('_', '-', params$date)`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: kable
params:
  site:        "rr"
  date:        "2022_08_10"
  ecp_path:    "X:/legacy/gdrive/saltmarsh_UAS_native/In Situ Data Collection/JoshSurveyPoints_AllSites_One_Sheet.xlsx"
  base_output: "E:/uas_scratch/lidar/rr/2022_08_10"
  eval_dir:    "lidar/output/rr_2022_08_10"
  tolerances:  [0.10, 0.20]
  ecp_types:   ["EVP"]
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
library(here)
library(ggplot2)
setwd(here::here())

invisible(lapply(list.files("R/", pattern = "\\.[Rr]$",
                            full.names = TRUE), source))

# -- Load ECPs ----------------------------------------------------------------
site_ecp <- load_ecp(params$ecp_path, site = params$site)
if (length(params$ecp_types) > 0L) {
   site_ecp <- site_ecp[
      tolower(site_ecp$type) %in% tolower(params$ecp_types), ,
      drop = FALSE
   ]
}

# -- List DTMs ----------------------------------------------------------------
dtm_dir   <- file.path(params$base_output, "zzzraster")
dtm_paths <- list.files(dtm_dir, pattern = "^csf_.*\\.tif$",
                        full.names = TRUE)
if (length(dtm_paths) == 0L) {
   stop("No DTMs found in: ", dtm_dir,
        "\nRun lidar/02.R first.")
}

# -- Sample DTMs at ECP locations (cache hit in normal use) -------------------
all_elevations <- do.call(rbind, lapply(dtm_paths, function(d) {
   elev     <- sample_dtm(d, site_ecp)
   elev$dtm <- sub("\\.tif$", "", basename(d), ignore.case = TRUE)
   elev
}))


# Subset to just ecp_types
all_elevations <- all_elevations[all_elevations$type %in% params$ecp_types, ]




# -- Evaluate each DTM --------------------------------------------------------
results       <- list()
all_summaries <- list()
tolerances    <- params$tolerances

for (d in dtm_paths) {
   stem       <- sub("\\.tif$", "", basename(d), ignore.case = TRUE)
   dtm_elev   <- all_elevations[all_elevations$dtm == stem, ]
   result     <- evaluate_dtm(dtm_elev, tolerances = tolerances)
   summary_d  <- result$summary
   summary_d$dtm <- stem
   results[[stem]]      <- result
   all_summaries[[stem]] <- summary_d
}

# -- Combined summary ---------------------------------------------------------
all_summary <- do.call(rbind, all_summaries)
all_summary <- all_summary[, c("dtm", "group",
                               setdiff(colnames(all_summary),
                                       c("dtm", "group")))]

# -- Write summary CSV --------------------------------------------------------
dir.create(params$eval_dir, recursive = TRUE, showWarnings = FALSE)
csv_path <- file.path(params$eval_dir, "dtm_eval_summary.csv")
write.csv(all_summary, csv_path, row.names = FALSE)

# -- Report helpers -----------------------------------------------------------
dtm_stems    <- unique(all_summary$dtm)
pct_cols     <- grep("^pct_within", colnames(all_summary), value = TRUE)
overall_rows <- all_summary[all_summary$group == "overall", ]
flagged_bias <- overall_rows[
   !is.na(overall_rows$bias_p) & overall_rows$bias_p < 0.05, "dtm"
]
rnd <- function(x, d = 3) round(as.numeric(x), d)
```

## Overview

```{r overview, results = "asis"}
n_ecp_overall <- overall_rows$n[1]
n_dtms        <- length(dtm_stems)

cat("**Site:**", params$site, " | ",
    "**Date:**", gsub("_", "-", params$date), " | ",
    "**DTMs evaluated:**", n_dtms, " | ",
    "**ECPs (overall):**", n_ecp_overall)
```

```{r ecp_class_counts}
class_rows <- all_summary[all_summary$dtm == dtm_stems[1] &
                             all_summary$group != "overall", ]
class_tbl <- class_rows[c("group", "n")]

colnames(class_tbl) <- c("Vegetation class", "n")
knitr::kable(class_tbl[order(-class_tbl$n), ],
             row.names = FALSE,
             caption = "ECP counts by class (Logger Array excluded).")
```

```{r bias_flag, results = "asis"}
if (length(flagged_bias) > 0) {
   cat("> **Note:** The following DTMs have a mean vertical bias",
       "statistically distinguishable from zero (t-test p < 0.05):",
       paste(flagged_bias, collapse = ", "),
       ". Check the residual plots for systematic patterns.\n")
} else {
   cat("> Mean vertical bias is not statistically distinguishable",
       "from zero in any DTM (t-test p ≥ 0.05 for all).\n")
}
```

---


### Overall RMSE

```{r overall_rank}
show_cols <- c("dtm", "n", "mean_bias", "rmse", "mae", pct_cols)
show_cols <- show_cols[show_cols %in% colnames(overall_rows)]
rank_tbl  <- overall_rows[order(overall_rows$rmse), show_cols]
rank_tbl$mean_bias <- rnd(rank_tbl$mean_bias, 3)
rank_tbl$rmse      <- rnd(rank_tbl$rmse, 3)
rank_tbl$mae       <- rnd(rank_tbl$mae, 3)
for (pc in pct_cols) rank_tbl[[pc]] <- rnd(rank_tbl[[pc]], 1)
colnames(rank_tbl)[colnames(rank_tbl) == "dtm"]       <- "DTM"
colnames(rank_tbl)[colnames(rank_tbl) == "mean_bias"] <- "Bias (m)"
colnames(rank_tbl)[colnames(rank_tbl) == "rmse"]      <- "RMSE (m)"
colnames(rank_tbl)[colnames(rank_tbl) == "mae"]       <- "MAE (m)"
knitr::kable(rank_tbl, row.names = FALSE)
```

---

## Full metrics table

All metric columns, all DTMs and classes.

```{r full_table}
show <- all_summary
levs <- unique(show$group) |> sort()
levs <- levs <- c("overall", levs[!levs == "overall"])
show$group <- ordered(show$group, levels = levs)

show <- dplyr::arrange(show, group)
num_cols <- c("mean_bias", "median_bias", "rmse", "mae", "mad",
              "max_abs", "q25_abs", "q75_abs", "q95_abs",
              "slope", "bias_p")
for (col in num_cols[num_cols %in% colnames(show)]) {
   show[[col]] <- rnd(show[[col]], 3)
}
for (pc in pct_cols) show[[pc]] <- rnd(show[[pc]], 1)
knitr::kable(show, row.names = FALSE)
```

---

## Cross-DTM comparison

Predicted vs. observed elevation across all DTMs on a common scale,
colored by vegetation class.
Points near the 1:1 line indicate accurate DTM estimation;
scatter above the line (predicted > observed) is typical where
tall vegetation was not fully filtered.

### Standard plot

```{r cross_dtm_plot, fig.width = 10, fig.height = 10, out.width= "100%"}
vhc_floor  <- 0.005   # m; replaces 0 so log scale is defined
plot_data  <- all_elevations[all_elevations$type %in% params$ecp_types, ]


color_by <- "subclass"

stopifnot(color_by %in% c("veg_height_m", "subclass", NA))

if (!color_by %in% colnames(plot_data)) {
   warning("plot data (elevations) doesn't have a \"", color_by, "\" column")
   color_by <- NA
}


if (color_by == "veg_height") {
   plot_data$vhc_plot <- pmax(as.numeric(plot_data$veg_height_m),
                              vhc_floor)
   p <- ggplot(plot_data,
               aes(x = elevation, y = predicted, color = vhc_plot)) +
      scale_color_viridis_c(
         name     = "Veg height (m)",
         trans    = "log10",
         na.value = "grey50",
         breaks   = c(0.001, 0.01, 0.1, 1),
         labels   = c("0.01", "0.01", "0.1", "1")
      )
} else if (color_by == "subclass") {
   p <- ggplot(plot_data,
               aes(x = elevation, y = predicted, color = subclass))
} else {
   p <- ggplot(all_elevations,
               aes(x = elevation, y = predicted))
}


# Standard to all plots and color schemes
p <- p  +
   geom_point(size = 1.2, alpha = 0.7) +
   facet_wrap(~ dtm, labeller = label_wrap_gen(20)) +
   geom_abline(slope = 1, intercept = 0,
               linetype = "dashed", color = "grey40") +
   geom_abline(slope = 1, intercept = 0.1,
               linetype = "dotted", color = "grey60") +
   #  coord_fixed() +
   labs(x = "Observed elevation (m)", y = "DTM elevation (m)") +
   theme_minimal()

# Render plot (without vertical lines)
print(p)
```

# With vegetation height indicated

Vertical lines stretch from the RTK field surveyed elevation to the 
top of the vegetation as recordered in the field.


```{r cross_dtm_plot_lines, fig.width = 10, fig.height = 10, out.width= "100%"}


# Render plot with height lines
if ("veg_height_m" %in% names(plot_data)) {
   p2 <- p + geom_segment(aes(x = elevation, xend = elevation,
                              y = elevation,
                              yend = elevation + veg_height_m,
                              color = !!sym(color_by)))
   print(p2)
}

```

---

## Per-DTM plots {.tabset}

Five diagnostic plots per DTM.
**Q-Q plot:** a straight line means residuals are roughly normal,
which validates the t-test used for `bias_p`;
heavy tails suggest using a sign test instead.

```{r per_dtm_plots, results = "asis"}
plot_labels <- c(
   pred_vs_obs           = "Predicted vs observed",
   residual_hist         = "Residual histogram",
   residual_map          = "Residual map",
   residual_vs_elevation = "Residual vs elevation",
   qq_residuals          = "Q-Q plot"
)

for (stem in dtm_stems) {
   cat("\n### ", stem, "\n\n", sep = "")
   res <- results[[stem]]
   for (pname in names(plot_labels)) {
      if (pname %in% names(res$plots)) {
         cat("**", plot_labels[[pname]], "**\n\n", sep = "")
         print(res$plots[[pname]])
         cat("\n\n")
      }
   }
}
```
