---
title: "Ground Elevation Source Comparison — `r params$site`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: kable
params:
  site:       "oth"
  sources:  NULL
  ecp_path:   NULL
  output_dir: NULL
  tolerances: [0.10, 0.20]
  ecp_types:  ["EVP"]
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
library(here)
library(ggplot2)
library(dplyr)
library(pathtools)
setwd(here::here())

invisible(lapply(list.files("R/", pattern = "\\.[Rr]$",
                            full.names = TRUE), source))
set_path_scheme("lidar/data/paths.yml")

# Unpack params to local variables for convenience.
site       <- params$site
sources    <- params$sources
ecp_path   <- params$ecp_path %||% get_path("ecp_path")
output_dir <- params$output_dir %||% get_path("ground_comparison_report",
                                              site = site)
tolerances <- params$tolerances
ecp_types  <- params$ecp_types

# Default sources for standalone knitting (oth values).
if (is.null(sources) || length(sources) == 0L) {
   if (!site == "oth")
      stop("Default sources are for site oth")
   sources <- c(
      may_14_ortho   = get_path("photo_dem", site = site,
                                source = "may_14_ortho"),
      may_02_ortho   = get_path("photo_dem", site = site,
                                source = "may_02_ortho"),
      massgis_346737 = get_path("massgis_tile", tile = "19TCH346737"),
      massgis_346735 = get_path("massgis_tile", tile = "19TCH346735")
   )
}


if (is.null(sources) || length(sources) == 0L) {
   stop("sources param is required — pass a named character vector ",
        "of label -> raster path.")
}

found   <- file.exists(sources)
missing <- sources[!found]
sources <- sources[found]
if (length(sources) == 0L) stop("No source files found.")

# -- Load ECPs ----------------------------------------------------------------
site_ecp <- load_ecp(ecp_path, site = site)
if (length(ecp_types) > 0L) {
   site_ecp <- site_ecp[
      tolower(site_ecp$type) %in% tolower(ecp_types), ,
      drop = FALSE
   ]
}

# -- Sample each source at ECP locations (skip-if-exists cache) ---------------
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

all_elevations <- do.call(rbind, lapply(names(sources), function(label) {
   csv_path <- file.path(output_dir, paste0(label, "_ecp.csv"))
   elev     <- sample_dtm(sources[[label]], site_ecp,
                          output_csv = csv_path)
   elev$source <- label
   elev
}))

# -- Evaluate each source -----------------------------------------------------
results       <- list()
all_summaries <- list()

for (label in names(sources)) {
   elev             <- all_elevations[all_elevations$source == label, ]
   result           <- evaluate_dtm(elev, tolerances = tolerances)
   summary_l        <- result$summary
   summary_l$source <- label
   results[[label]]       <- result
   all_summaries[[label]] <- summary_l
}

all_summary <- do.call(rbind, all_summaries)
all_summary <- all_summary[, c("source", "group",
                               setdiff(colnames(all_summary),
                                       c("source", "group")))]

# -- Write summary CSV --------------------------------------------------------
csv_out <- file.path(output_dir, "ground_source_comparison.csv")
write.csv(all_summary, csv_out, row.names = FALSE)

# -- Report helpers -----------------------------------------------------------
pct_cols     <- grep("^pct_within", colnames(all_summary), value = TRUE)
overall_rows <- all_summary[all_summary$group == "overall", ]
rnd <- function(x, d = 3) round(as.numeric(x), d)
```

## Overview

```{r overview, results = "asis"}
cat("**Site:**", site, " | ",
    "**ECPs:**", overall_rows$n[1],
    paste0("(type = ", paste(ecp_types, collapse = ", "), ")"),
    " | ",
    "**Sources evaluated:**", length(sources))

if (length(missing) > 0L) {
   cat("\n\n> **Missing sources (skipped):** ",
       paste(names(missing), collapse = ", "))
}
```

---

## Bias and RMSE by source

```{r rmse_table}
show_cols <- c("source", "n", "mean_bias", "rmse", "mae", pct_cols)
show_cols <- show_cols[show_cols %in% colnames(overall_rows)]
rank_tbl  <- overall_rows[order(overall_rows$rmse), show_cols]
rank_tbl$mean_bias <- rnd(rank_tbl$mean_bias, 3)
rank_tbl$rmse      <- rnd(rank_tbl$rmse, 3)
rank_tbl$mae       <- rnd(rank_tbl$mae, 3)
for (pc in pct_cols) rank_tbl[[pc]] <- rnd(rank_tbl[[pc]], 1)
colnames(rank_tbl)[colnames(rank_tbl) == "mean_bias"] <- "Bias (m)"
colnames(rank_tbl)[colnames(rank_tbl) == "rmse"]      <- "RMSE (m)"
colnames(rank_tbl)[colnames(rank_tbl) == "mae"]       <- "MAE (m)"
knitr::kable(rank_tbl, row.names = FALSE)
```

A consistent positive `mean_bias` across sources that share a common
processing lineage (e.g. all project lidar and photogrammetry) but not
in external reference data (MassGIS) points to a datum or reprojection
issue rather than a ground-finding problem.

```{r bias_plot, fig.height = 4}
plot_overall <- overall_rows
plot_overall$source <- factor(
   plot_overall$source,
   levels = plot_overall$source[order(plot_overall$mean_bias)]
)

ggplot(plot_overall, aes(x = source, y = mean_bias,
                         fill = mean_bias > 0)) +
   geom_col(show.legend = FALSE) +
   geom_errorbar(aes(ymin = mean_bias - rmse,
                     ymax = mean_bias + rmse),
                 width = 0.25, color = "grey30") +
   geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
   coord_flip() +
   scale_fill_manual(values = c("TRUE" = "#d73027",
                                "FALSE" = "#4575b4")) +
   labs(x = NULL, y = "Mean bias (m)",
        caption = "Error bars span +/- RMSE.") +
   theme_minimal()
```

---

## Full metrics table

All metric columns, all sources and ECP classes.

```{r full_table}
show  <- all_summary
levs  <- sort(unique(show$group))
levs  <- c("overall", levs[levs != "overall"])
show$group <- ordered(show$group, levels = levs)
show  <- dplyr::arrange(show, group, mean_bias)
num_cols <- c("mean_bias", "median_bias", "rmse", "mae", "mad",
              "max_abs", "q25_abs", "q75_abs", "q95_abs",
              "slope", "bias_p")
for (col in num_cols[num_cols %in% colnames(show)]) {
   show[[col]] <- rnd(show[[col]], 3)
}
for (pc in pct_cols) show[[pc]] <- rnd(show[[pc]], 1)
knitr::kable(show, row.names = FALSE)
```

---

## Predicted vs. observed by source

```{r cross_source_plot, fig.width = 10, fig.height = 10, out.width = "100%"}
ggplot(all_elevations,
       aes(x = elevation, y = predicted, color = subclass)) +
   geom_point(size = 1.2, alpha = 0.7) +
   facet_wrap(~ source, labeller = label_wrap_gen(20)) +
   geom_abline(slope = 1, intercept = 0,
               linetype = "dashed", color = "grey40") +
   labs(x = "Observed elevation (m)", y = "Predicted elevation (m)",
        color = "Subclass") +
   theme_minimal()
```

---

## Bias correction

The results above suggest a systematic bias in the UAS-based data
sources, possibly a datum shift or other projection issue.
This apparent bias swamps other sources of error and makes comparison
difficult.
The plots and tables here use data with the per-source bias removed.

```{r bias_adjustment}
estimate_bias <- function(x, y) {
   # Return NA when there are no non-NA pairs (e.g. raster doesn't
   # overlap the ECP extent for this source).
   if (sum(!is.na(x) & !is.na(y)) == 0L) return(NA_real_)
   fit <- lm(y ~ 1, offset = x)
   fit$coefficients[1] |> as.numeric()
}

unbiased <- all_elevations
for (i in seq_along(sources)) {
   sv   <- all_elevations$source == names(sources[i])
   bias <- estimate_bias(x = all_elevations$elevation[sv],
                         y = all_elevations$predicted[sv])
   if (!is.na(bias)) {
      unbiased$predicted[sv] <- all_elevations$predicted[sv] - bias
   }
}

unbiased$residual <- unbiased$predicted - unbiased$elevation

unbiased_results    <- list()
unbiased_summaries  <- list()

for (label in names(sources)) {
   elev             <- unbiased[unbiased$source == label, ]
   result           <- evaluate_dtm(elev, tolerances = tolerances,
                                    skip_plots = TRUE)
   summary_l        <- result$summary
   summary_l$source <- label
   unbiased_results[[label]]    <- result
   unbiased_summaries[[label]]  <- summary_l
}

unbiased_summary <- do.call(rbind, unbiased_summaries)
unbiased_summary <- unbiased_summary[
   , c("source", "group",
       setdiff(colnames(all_summary), c("source", "group")))
]
uo_rows          <- unbiased_summary$group == "overall"
unbiased_overall <- unbiased_summary[uo_rows, , drop = FALSE]
```

### Summary stats by source after bias correction

```{r unbiased_overall_table}
show_cols <- c("source", "n", "mean_bias", "rmse", "mae", pct_cols)
show_cols <- show_cols[show_cols %in% colnames(unbiased_overall)]
rank_tbl  <- unbiased_overall[order(unbiased_overall$rmse), show_cols]
rank_tbl$mean_bias <- rnd(rank_tbl$mean_bias, 3)
rank_tbl$rmse      <- rnd(rank_tbl$rmse, 3)
rank_tbl$mae       <- rnd(rank_tbl$mae, 3)
for (pc in pct_cols) rank_tbl[[pc]] <- rnd(rank_tbl[[pc]], 1)
colnames(rank_tbl)[colnames(rank_tbl) == "mean_bias"] <- "Bias (m)"
colnames(rank_tbl)[colnames(rank_tbl) == "rmse"]      <- "RMSE (m)"
colnames(rank_tbl)[colnames(rank_tbl) == "mae"]       <- "MAE (m)"
knitr::kable(rank_tbl, row.names = FALSE)
```

### DTM/DEM elevation vs ground truth after bias correction

```{r unbiased_plots, fig.width = 10, fig.height = 10, out.width = "100%"}
ggplot(unbiased,
       aes(x = elevation, y = predicted, color = subclass)) +
   geom_point(size = 1.2, alpha = 0.7) +
   facet_wrap(~ source, labeller = label_wrap_gen(20)) +
   geom_abline(slope = 1, intercept = 0,
               linetype = "dashed", color = "grey40") +
   labs(x = "Observed elevation (m)",
        y = "Predicted elevation - bias (m)",
        color = "Subclass") +
   theme_minimal()
```

---

## Per-source diagnostic plots {.tabset}

Using the original, biased data.

```{r per_source_plots, results = "asis"}
plot_labels <- c(
   pred_vs_obs           = "Predicted vs observed",
   residual_hist         = "Residual histogram",
   residual_map          = "Residual map",
   residual_vs_elevation = "Residual vs elevation",
   qq_residuals          = "Q-Q plot"
)

for (label in names(sources)) {
   cat("\n### ", label, "\n\n", sep = "")
   res <- results[[label]]
   for (pname in names(plot_labels)) {
      if (pname %in% names(res$plots)) {
         cat("**", plot_labels[[pname]], "**\n\n", sep = "")
         print(res$plots[[pname]])
         cat("\n\n")
      }
   }
}
```
